Mining Interaction Log Data in a Creative Arts Practice using Transition Matrices

### 2015-09-24 flux-entropy-paper/flux-entropy-paper.tex
cut some comments
final updates! uploading now.

%\title{Mining interaction log data in long-term practice-led research}

@@

In all of our 95 interaction sessions, every touch interaction 
(touches begin, touches move, touches end) was time-stamped and
logged to disk for later analysis. These sessions included {\em rehearsals} without an
audience, as well as {\em concerts}, in front of a live audience, where 
performative goals were paramount. They also included a number of {\em
  recording} sessions
for the purposes of formally-structured HCI experiments. In the recording
sessions, the performers were also interviewed and surveyed-- something which would have 
been disruptive and inappropriate for the concerts. Survey and interview data were not part of the
corpus. 
Some of the sessions were
iPad-only and others used iPads together with other acoustic
percussion instruments. Although the majority of the sessions were
free-form improvisations, a number of rehearsals and performances
involved composed pieces where the musicians read and performed
percussion notation using the apps.
Some of these session contexts are are shown
Figure~\ref{fig:three-performance-contexts} and other descriptive
statistics about the corpus are shown in Table \ref{corpus-table}.
In our corpus, the median length of sessions is 7m26s and the median
number of participants per ensemble is four. Each session record is
labelled with metadata about its \emph{context} (concert,
rehearsal, recording), its \emph{type} (improvisation or composition),
and its \emph{instrumentation} (iPads-only, or iPads with acoustic
percussion). The distribution of recorded sessions by context and type
is shown in Figure \ref{fig:count-data}.

% It is a feature of creative expression in the arts that highly skilled
% practitioners exploit simple tools to produce works of beauty whether
% it be static paintings or dynamic performance. 
% In previous work~\cite{Martin:2014cr} researchers have shown that
% trained percussionists can exploit a very simple iPad interface to
% discover new categories of musical gestures.
% In that paper, iPad apps
% were studied where tapping was mapped to produce produced short
% sounds, while swiping produced long sounds with the volume
% proportional to the velocity of the moving touch point. 
% It turned out that trained percussionists working together in an
% ensemble were able to discover and repeatedly make use of five macro
% gestures made up of taps and swipes: fast taps, slow taps, random
% taps, short swipes and swirls. This process of discovery of
% interesting musical gestures from a simple interface is analogous to
% what percussionists do with acoustic instruments -- drums and bowls
% and bells are often tapped, stroked and hit in novel ways to
% communicate meaning and emotion through music~\cite{Schick:2006fk}.

@@

In our corpus of iPad musical performances, logs of low-level touch-data record each performer's micro-gestures. 
Because it has been
established that musical performances on touch-screens define  
vocabularies of continuous macro gestures that unfold over
several seconds of touch interaction~\cite{Martin:2014cr}, and that
gestural performances can be divided into sequences of such states~\cite{Pressing:1988uo}, these logs
 can be used identify a sequence of such states for each performer.
If a performers' touch interactions can be classified as belonging
to one of $m$ distinct states, then, over the course of a session, each
participant $X$ generates a length $N$ sequence of these interaction
states $X_n$ ($n = 1, \ldots, N$) which can be modelled as a Markov
process.
If there is an (automated) classification process for identifying each
participant's state sequence from their interaction data, then we can
estimate the $m \times m$ transition matrix (TM), $P$, which characterises
this process by setting each entry, $p_{ij}$, to be the proportion of times
state $j$ follows state $i$ in the sequence. The transition activity
of the whole ensemble can be calculated by averaging the TM for each
performer. The matrix can be normalised so that the sum of all
elements is equal to one.

@@

%HCI researchers are increasingly concerned with interactions that
%happen in-the-wild, where the creative collaborations of many users
%may be recorded, but where conducting surveys or interviews is
%innapropriate. We suggest that analysing interaction-state transitions
%using matrices and simple measures can lead to new insights into HCI
%interactions outside the lab.

\pagebreak
\bibliographystyle{SIGCHI-Reference-Format}
\bibliography{references}

>>>>

@@

In all of our 95 interaction sessions, every touch interaction 
(touches begin, touches move, touches end) was time-stamped and
logged to disk for later analysis. These sessions included {\em rehearsals} without an
audience, as well as {\em concerts}, in front of a live audience, where 
performative goals were paramount. They also included a number of {\em
  recording} sessions
for the purposes of formally-structured HCI experiments. In the recording
sessions, the performers were also interviewed and surveyed-- something which would have 
been disruptive and inappropriate for the concerts. Survey and interview data were not part of the
corpus. 
Some of the sessions were
iPad-only and others used iPads together with other acoustic
percussion instruments. Although the majority of the sessions were
free-form improvisations, a number of rehearsals and performances
involved composed pieces where the musicians read and performed
percussion notation using the apps.
Some of these session contexts are are shown in 
Figure~\ref{fig:three-performance-contexts} and other descriptive
statistics about the corpus are shown in Table \ref{corpus-table}.
In our corpus, the median length of sessions is 7m26s and the median
number of participants per ensemble is four. Each session record is
labelled with metadata about its \emph{context} (concert,
rehearsal, recording), its \emph{type} (improvisation or composition),
and its \emph{instrumentation} (iPads-only, or iPads with acoustic
percussion). The distribution of recorded sessions by context and type
is shown in Figure \ref{fig:count-data}.

@@

In our corpus of iPad musical performances, logs of low-level touch-data record each performer's micro-gestures. 
Because it has been
established that musical performances on touch-screens define  
vocabularies of continuous macro gestures that unfold over
several seconds of touch interaction~\cite{Martin:2014cr}, and that
gestural performances can be divided into sequences of such states~\cite{Pressing:1988uo}, these logs
 can be used to identify a sequence of such states for each performer.
If a performers' touch interactions can be classified as belonging
to one of $m$ distinct states, then, over the course of a session, each
participant $X$ generates a length $N$ sequence of these interaction
states $X_n$ ($n = 1, \ldots, N$) which can be modelled as a Markov
process.
If there is an (automated) classification process for identifying each
participant's state sequence from their interaction data, then we can
estimate the $m \times m$ transition matrix (TM), $P$, which characterises
this process by setting each entry, $p_{ij}$, to be the proportion of times
state $j$ follows state $i$ in the sequence. The transition activity
of the whole ensemble can be calculated by averaging the TM for each
performer. The matrix can be normalised so that the sum of all
elements is equal to one.

@@

\break
\bibliographystyle{SIGCHI-Reference-Format}
\bibliography{references}
